Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees

### 2023-03-02 script/manuscript_MEE.Rmd
many edits from JEO PJC, and initiation of table

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
    |  $^4$ Australia post fire project, ----, ----, ----, AU  
    |  Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^,$$^4$, Emily J. Woodworth$^1$$^,$$^2$, 
abstract:  |  
  |  
  1) DNA Barcoding has been successful for the rapid analysis of ecological assemblages. Although barcoding in the plant kingdom barcoding has been more difficult than others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy for this barcoding system we compared the data to museum species, observation studies, and species distribution modelling to identify likely candidate species. 
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches. 
  6) Understanding plants in ecological contexts and understandings of their synecology.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

The inability to reliably identify plants to terminal taxon can limit our understanding of ecosystem function and interactions (@bortolus2008error). This is especially true for genera where identification down to species can be useful as specific bioindicators; defining ecological and behavioral properties  (e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO).In these instances the lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and their interactions with other species. This can be further complicated by the fact that the identification of organisms to terminal taxon is also often mired by lack of diagnostic characters(e.g. flowers, fruits, roots or combinations thereof), an increasing lack of taxonomic experts (@hebert2003biological) and the presence of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Taxonomic verification can also be limited by the fact that revisiting field sites to identify material using morphological or chemical approaches, can be resource intensive and often does not work. The current methods to ameliorate this situation include: ignoring these ecologically relevant levels of detail, revisiting plots as diagnostic material becomes temporally available, seeking the assistance from taxonomic specialists, or the use of barcoding using molecular techniques (CITE)

  Recently molecular barcoding (the identification of a sample from a single organism *e.g.* a piece of leaf), or metabarcoding (the identification of a sample containing a mix of organisms *e.g.* soil), have shown considerable promise in many taxa (@ruppert2019past). For plants the success is a little more mix, with the identification of certain clades using barcoding being quite successful (@kress2017plant), however for many other clades the results have been more elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), while metabarcoding has incurred additional challenges for the currently available barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges for the utilization of the high copy number barcodes (e.g. ITS2, *rbcL*, *matK*, *trnH-psbA*) include their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant).

  Currently the largest plant systematic endeavor ever undertaken,by the Royal Botanic Gardens Kew, the Plant and Fungal Tree of Life (PAFTOL) is approaching completion (@baker2021PAFTOL).  This data set will contain hybridization capture (Hyb-Seq) data from at least one species in each genus of the plant kingdom,14,000 represented species, using the popular Angiosperms353 (A353) probes, which includes 353 single-copy orthologous loci, (@baker2021PAFTOL, @johnson2019universal). These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. The A353 probes are currently being used in many other plant phylogenetic studies increasing the sampling depth of many clades  (@baker2021exploring). Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more species upon it’s intended completion, slated for 2030. Similar projects such as the ‘Darwin Tree of Life’ which will sequence all described taxa in Britain and Ireland, seek to sequence high amounts of genomes in geographic regions will contribute data sets applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diverse array of questions relevant to theoretical and applied ecology (@kress2017plant, @hollingsworth2016telling). However, the application of metabarcoding still face challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data.

@@

  To increase the quality of metabarcoding results in plants, we are proposing reducing the number of possible candidate species by generating a user specific databases relevant to the region of study and the ecological characteristics of interest (@bell2022plants). To achieve this goal, we first create a regional list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. For these candidate species, a modelling approach, such as logistic regression, may be used to identify taxa which warrant further exploration e.g. determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where in the study area might suitable habitats. This approach has the benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This approach can significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

@@

@@

## 2.1 Study System & Field Work

Observations and bee sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information), characterized by high-montane/subalpine Parkland vegetation communities. Pollinator observations of *Bombus* Latreille spp. (Apidae Latreille) were conducted from May 29^th^ - July 23^rd^ of 2015 in six study sites. Observations of *Bombus* foraging took place for one hour at each field site in three 50m transects, each transect was in a major vegetation type (dry, and wet meadows, and Aspen forest), where all flowers abundances were estimated and placed into binds of abundance. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

## 2.2 | Floral Visitation

*... does this need more info and warrant a section independent of the above?*

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner.

@@

  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2^ of pollen was placed onto a ~4mm^2^ fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish to prevent oxidation; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

We generate a short list of potential candidate species we downloaded from the Botanical Information and Ecology Network ‘BIEN’ (@bien2022) all records adjacent to the field sites to develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study area. To reduce the list of species to include in the genomic sequence databases, we then generated Species Distribution Models (SDMs) for these taxa to predict their distribution throughout the study area.

In order to minimize the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. To account for the stochasticity of botanical collecting and offset the number of records associated with the research station, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

We used all occurrence records from BIEN (n = 23,919) within a 50km border of the Omernik level 3 ecoregion, which includes the study area *(No. 21 “Southern Rockies”)* to construct the species distribution model  (@omernik1987ecoregions). These records were copied into two, initially identical, sets, one for generating machine learning models (ML; Random Forest, and Boosted Regression Tree’s), and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM) (@barbet2012selecting). Ensembled predictions have been shown to outperform their constituent models, on average, and to reduce the ecological signal to the analytical noise of individual runs (@araujo2007ensemble). No single method of producing SDMs has been shown to universally outperform others when faced with a large and diverse number of applications, in our case a great number of species with differing biologies and ecologies (@elith2006novel, @qiao2015no). In the spirit of these findings, multiple families of models, which can be generated together as they have similar requirements regarding the number and ratios of Presence to Absence records were ensembled together (@barbet2012selecting).

@@

To predict the potential distribution of each species we used 26 environmental variables at 30m resolution, six related to climate, five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). **These publicly available data sets, were selected as they ...** . For linear regression models these predictors underwent both *vifstep* (theta = 10, max observations = 12,500) and *vifcor* (theta = 0.7, max observations = 12,500) to detect highly correlated variables, and collinear features were removed leaving 16 variables (@usdm2014).

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 ecoregions if enough records were not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS data set (@iler2021conceptual, @tran2019cloud) from 2000-2017 were analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3^rd^ quantile for melt, and the 1^st^ quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered, we used the initiation and cessation dates, respectively, as effective start and ends of flowering. These estimates were compared to groundtruthed data from a long term experiment 1974-2012, and the field data from 2015, using kendall's tau.

@@

Using five years (2015-2020) of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples collected from one site within the Gunnison Basin River Drainage and one individual collected from another more distal population. In addition we included a congener - or a species from a closely related genus to serve as an outgroup for all 12 taxa. In addition we sequenced another 15 taxa commonly visited by Bombus workers, based on the abundances, and immediate access to plant tissue, in the aforementioned data set  (*APPENDIX 4*). Plant collections were identified typically using a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

@@

Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°C for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

To precipitate the DNA, chilled Isopropyl alcohol & 3 mM Sodium acetate (5:1) equivalent to $\frac{2}{3}$ of the volume of sample were added, with 1 hour of chilling at -20°C, followed by 10 minutes of centrifuging at 13,000 rpm. The supernatant was pipetted to a new 1.5 ml centrifuge tube, and 70% EtOH (400 µL) were added before chilling at -20°C for 20 minutes followed by centrifugation at 13,000 rpm for 10 minutes. Both tubes were then washed with  75% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded, then washed with 95% EtOH (400 µL) , inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded. Pellets were dried at room temperature overnight before resuspension in nuclease free H~2~O.  Extractions were assessed using a Nanodrop 2000 (Thermo Fisher Scientific) and Qubit fluorometer (Thermo Fisher Scientific). DNA extracts were then cleaned using 2:1 v./v. Sera-Mag beads (Cytiva, Little Chalfont, UK) to solute ratio following the manufacturer's protocol, eluted in 0.5x TE, and the eluent allowed to reduce by half volume in ambient conditions. DNA was quantified using a Qubit fluorometer.

@@

In order to precisely classify the contents of each corbiculae load the sequences classified by molecular methods were compared with the fieldwork which at a very fine resolution, recorded the presence and absence of species and their duration of flowering, and was interpreted ala the computer derived temporal and spatial data sets. The quantitative counts of grains from microscopy, were combined with the semi-quantitative sequencing results, to estimate the abundance of each identified species in each corbiculae load.

To reclassify the sequence reads, these data were combined with the flora observation data, and mapped by genus. If more than one species in the genus was flowering at that time and site, than the reads were split evenly between the taxa. For sequence data which did not match at the genus level, a user subjectively scored them based on the species composition and phenological activity at each plot, the Queen interaction data, and pollen assignments. To estimate the abundance of each of these species in the corbiculae loads, these data were combined with the microscopy data. For each morphotype detected in pollen, and each classified sequence read which was not detected via microscopy, they were given a value of 0.5% to indicate there trace presences. When more than a single species belonged to a morphotype group in a single sample, the quantitative values from the morphological work were multiplied by the relative sequence abundance of each species in the load. All final compositions were standardized to a sum of 100%, by adding or subtracting the differences (induced by classifying records as 'trace') to all species with abundances > 1%.

>>>>

--- 
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging Patterns of Queen Bumble Bees
author:  |
    | Reed Clark Benkendorf$^1$$^,$$^2$^[Correspondence: reedbenkendorf2021@u.northwestern.edu], Jane E. Ogilive$^3$$^, Emily J. Woodworth$^1$$^,$$^2$, 
    |  Sophie Taddeo$^1$$^,$$^2$, Paul J. CaraDonna$^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant$^1$$^,$$^2$  
    |  $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022, USA  
    |  $^2$Plant Biology and Conservation, Northwestern University, Evanston, Illinois 60208, USA  
    |  $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, Colorado 81224, USA
abstract:  |  
  |  
  1) DNA Barcoding has been successful for the rapid analysis of complex ecological assemblages. Barcoding in the plant kingdom barcoding has been more difficult than in others, and hence may begin to lag behind other kingdoms. 
  2) Here we test the utilization of Angiosperms 353 probes to barcode plant species found in corbiculae pollen loads collected from Queen Bumble Bees. 
  3) To verify the accuracy of this barcoding system we compared the barcoding data to museum plant specimens, observations of plant flowering and bee flower visitation in the field, and species distribution modelling to identify likely candidate plant species found within the corbiculae pollen loads.
  4) By utilizing Species distribution modelling we allow users to create a regionally appropriate sequence databases which may use increase the alignment algorithms minimizing need for large computational power, and run time. 
  5) We show that the Angiosperms 353 probes, developed for phylogenmoics, and which are currently being used in the largest ever plant systematic endeavor, offer significant promise to metagenomic approaches. 
  6) The DNA barcoding of bumble bee corbiculae pollen loads was most accurate when combined with knowledge of what plant species were flowering in the plant community when they were collected. Thus, supplementing DNA barcoding data with ecological context is most accurate and powerful.
keywords: |
  DNA baroding, metagenomics, hyb-seq, Angiosperms353, plant metagenomics, bombus queens, subalpine, species distribution model
output:
  pdf_document: default
  toc: no
  word_document: default
csl: "../citations/methods-in-ecology-and-evolution.csl"
bibliography: ../citations/citations.bib
link-citations: yes
fig_caption: yes
always_allow_html: yes
header-includes:
- \usepackage{endfloat}
- \usepackage{setspace}\doublespacing
- \usepackage{lineno}
- \linenumbers
---

@@

The inability to reliably identify plants to terminal taxon can limit our understanding of ecosystem function and interactions (@bortolus2008error). This is especially true for genera where identification down to species can be useful as specific bioindicators; defining ecological and behavioral properties  (e.g. different species of Sagebrush- *Artemisia* L., Willows - *Salix* L., and Sedges - *Carex* L.) (@Gage2013HistoricalRO).In these instances the lack of species level data can hinder our understanding of the breadth of habitat which some species occupy, and their interactions with other species. This can be further complicated by the fact that the identification of organisms to species is also often mired by lack of diagnostic characters(e.g. flowers, fruits, roots or combinations thereof), an increasing lack of taxonomic experts (@hebert2003biological) and the presence of cryptic species (@janzen2017nuclear, @oliver2009cryptic). Taxonomic verification can also be limited by the fact that revisiting field sites to identify material using morphological or chemical approaches, can be resource intensive and often ineffective. The current methods to ameliorate this situation include: (i) ignoring these ecologically relevant levels of detail, (ii) revisiting plots as diagnostic material becomes temporally available, (iii) seeking the assistance from taxonomic specialists, or (iv) the use of barcoding using molecular techniques (CITE).

  Currently the largest plant systematic endeavor ever undertaken,by the Royal Botanic Gardens Kew, the Plant and Fungal Tree of Life (PAFTOL) is approaching completion (@baker2021PAFTOL).  This data set will contain hybridization capture (Hyb-Seq) data from at least one species in each genus of the plant kingdom,14,000 represented species, using the popular Angiosperms353 (A353) probes, which includes 353 single-copy orthologous loci, (@baker2021PAFTOL, @johnson2019universal). These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. The A353 probes are currently being used in many other plant phylogenetic studies increasing the sampling depth of many clades  (@baker2021exploring).Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants, will contribute many more species by 2030 (@cheng2018tenkp). Similar projects such as the ‘Darwin Tree of Life’ which will sequence all described taxa in Britain and Ireland, seek to sequence high numbers of genomes in geographic regions will contribute data sets applicable to enormous spatial domains (@darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diverse array of questions relevant to theoretical and applied ecology (@kress2017plant, @hollingsworth2016telling). However, the application of metabarcoding still faces challenges relating to the enormity of the genomic data sets and the computational power required to process sequence data.

@@

  To increase the quality of metabarcoding results in plants, we are proposing reducing the number of possible candidate species by generating a user specific databases relevant to the region of study and the ecological characteristics of interest (@bell2022plants). To achieve this goal, we first create a regional list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. For these candidate species, a modelling approach, such as logistic regression, may be used to identify taxa which warrant further exploration e.g. determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models examine the ecological conditions associated with the known occurrence of a species to identify suitable habitats in the study area. This approach has the benefit of greatly reducing the size of a sequence database, which allows for the use of genomic data on personal computers. This approach can significantly reduce processing time, particularly as most next-generation sequence data is deposited as raw-sequence reads.

@@

@ogilvie2022shifting

@@

## 2.1 Case Study: Bee-Flower Observations and Pollen Load Collection

Bee and flower observations and bee corbiculae pollen collection was conducted around the Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). The area is characterized by subalpine meadow vegetation communities. Pollinator observations of Bombus Latreille spp. (Apidae Latreille) were conducted from May 29th – July 23rd of 2015 in six study sites as a part of a larger study (described in Ogilvie and CaraDonna 2022). Observations of Bombus foraging took place for one hour at each field site, with equal time spent searching for bee in the major vegetation types (dry, and wet meadows, and aspen forest). Corbiculae pollens loads were, non-lethally, collected from queens encountered by capturing them in an insect net and transferring them into a restraining device (“bee squeezer”, @kearns2001natural). We then collected a single pollen load (i.e., from one leg) from the bee and then released it. At weekly intervals at each site, we also recorded the abundances of flowers visited by bumble bees within belt transects spread over the three vegetation types (0.5 x 40 m transects in each vegetation type, 60 m^2^ total area per site).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non-accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner.

@@

  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2^ of pollen was placed onto a ~4mm^2^ fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with clear nail polish to prevent oxidation; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained a unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

We first generated a short list of potential candidate species. We downloaded from the Botanical Information and Ecology Network ‘BIEN’ (@bien2022) all records adjacent to the field sites to develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study area. To reduce the list of species to include in the genomic sequence databases, we then generated Species Distribution Models (SDMs) for these taxa to predict their distribution throughout the study area.

To minimize the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. To account for the stochasticity of botanical collecting and offset the number of records associated with the research station, this data set was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing data set. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

To determine which clades to include in the reference sequence database we used Species Distribution Modelling. We used all occurrence records from BIEN (n = 23,919) within a 50km border of the ecoregion, Omernik level 3, which includes the study area *(No. 21 “Southern Rockies”)* to construct the species distribution model  (@omernik1987ecoregions). These records were copied into two, initially identical, sets, one for generating machine learning models (ML; Random Forest, and Boosted Regression Tree’s), and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM) (@barbet2012selecting). Ensembled predictions have been shown to outperform their constituent models, on average, and to reduce the ecological signal to the analytical noise of individual runs (@araujo2007ensemble). No single method of producing SDMs has been shown to universally outperform others when faced with a large and diverse number of applications, in our case a great number of species with different biology and ecology (@elith2006novel, @qiao2015no). In the spirit of these findings, multiple families of models, which can be generated together as they have similar requirements regarding the number and ratios of Presence to Absence records were ensembled together (@barbet2012selecting).

@@

To predict the potential distribution of each species we used 26 environmental variables at 30m resolution, six related to climate, five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). These publicly available data sets, were selected as they pertain to a wide range of variables interacting with plant physiology. For linear regression models these predictors underwent both *vifstep* (theta = 10, max observations = 12,500) and *vifcor* (theta = 0.7, max observations = 12,500) to detect highly correlated variables, and collinear features were removed leaving 16 variables (@usdm2014).

@@

For assignment of reads to ecologically probabilistic species subsequent to BLAST, flowering time was used as a filter. To estimate the duration of dates in which plant species were flowering Weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 ecoregions if enough records were not found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS data set (@iler2021conceptual, @tran2019cloud) from 2000-2017 were analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in fall. Herbarium records after the 3^rd^ quantile for melt, and the 1^st^ quantile for snow cover of these metrics were removed. Species with > 10 records had their Weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered, we used the initiation and cessation dates, respectively, as effective start and ends of flowering. These estimates were compared to a long-term observational study of flowering phenology 1974-2012 (@caradonna2014shifts), and the floral abundance data from 2015, using Kendall's tau.

@@

Using five years (2015-2020) of observational data on *Bombus* queen interactions with flowering plants at these studies sites, we identified the plant taxa most frequently visited by queens across all years. We sequenced the 12 most visited taxa twice using samples collected from one site within the Gunnison Basin River Drainage and one individual collected from another more distal population. In addition we included a congener - or a species from a closely related genus to serve as an outgroup for all 12 taxa. We also sequenced another 15 taxa commonly visited by *Bombus* workers, based on the abundances, and immediate access to plant tissue, in the aforementioned data set  (*APPENDIX 4*). Plant collections were identified typically using a combination, of dichotomous keys and primary literature as required (@flora1993flora, @hitchcock2018flora,  @ackerfield2015flora, @lesica2012manual,  @cronquist1977intermountain, @allred2012flora, @jepson2022online, @mohlenbrock2002vascular).

@@

Pollen genomic DNA was extracted from corbiculae using a CTAB based protocol modified from Lahlamgiahi et al. and Guertler et al. (2014, 2014). A SDS extraction buffer (350µL , 100mM Tris-HCl, 50 mM EDTA, 50 mM NaCl, 10% SDS v/v., pH 7.5) was added followed by vortexing to allow dissolution of corbiculae. Pollen grains were then macerated with Kontes Pellet Pestles, and the tip of these washed with 130 µL of the SDS extraction buffer, samples were then incubated for 1 hour at 30°C. This was followed by the addition of 10% CTAB solution (450ul, of 20 mM Tris-Cl pH. 8.0, 1.4 M NaCl, 10 mM EDTA pH 7.5, 10% CTAB, 5% PVP, ~85% Deionized water) and RNAse (10 uL of 10 mg/mL) and samples were incubated for 40 minutes at 37°C, on a heat block (Multi-Blok, Thermo Fisher Scientific, Waltham Massachusetts) set to 40°C. After 20 minutes incubation, Proteinase K (15 µL of 20mg/ml) and  DTT (12.5 µL of 1M in water) were added, and the samples were further incubated at 60°C for 1 hour. Samples were then incubated overnight at 40°C. 500 µL of Phenol-Chloroform-Isoamyl alcohol (25:24:1) were added, vortexed, and centrifuged at 10,000 rpm for 10 minutes and the aqueous phase was pipetted to a 1.5 ml centrifuge tube.

To precipitate the DNA, chilled Isopropyl alcohol & 3 mM Sodium acetate (5:1) equivalent to $\frac{2}{3}$ of the volume of sample were added, with 1 hour of chilling at -20°C, followed by 10 minutes of centrifuging at 13,000 rpm. The supernatant was pipetted to a new 1.5 ml centrifuge tube, and 70% EtOH (400 µL) were added before chilling at -20°C for 20 minutes followed by centrifugation at 13,000 rpm for 10 minutes. Both tubes were then washed with  75% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded, then washed with 95% EtOH (400 µL), inverted, centrifuged at 13,000 rpm for 4 minutes, and the solution discarded. Pellets were dried at room temperature overnight before resuspension in nuclease free H~2~O.  Extractions were assessed using a Nanodrop 2000 (Thermo Fisher Scientific) and Qubit fluorometer (Thermo Fisher Scientific). DNA extracts were then cleaned using 2:1 v./v. Sera-Mag beads (Cytiva, Little Chalfont, UK) to solute ratio following the manufacturer's protocol, eluted in 0.5x TE, and the eluent allowed to reduce by half volume in ambient conditions. DNA was quantified using a Qubit fluorometer.

@@

To precisely classify the contents of each corbiculae load the sequences classified by molecular methods were compared with the fieldwork which at a very fine resolution, recorded the presence and absence of species and their duration of flowering, and was interpreted ala the computer derived temporal and spatial data sets. The quantitative counts of grains from microscopy, were combined with the semi-quantitative sequencing results, to estimate the abundance of each identified species in each corbiculae load.

To reclassify the sequence reads, these data were combined with the flora observation data, and mapped by genus. If more than one species in the genus was flowering at that time and site, than the reads were split evenly between the taxa. For sequence data which did not match at the genus level, a user subjectively scored them based on the species composition and phenological activity at each site, the queen interaction data, and pollen assignments. To estimate the abundance of each of these species in the corbiculae loads, these data were combined with the microscopy data. For each morphotype detected in pollen, and each classified sequence read which was not detected via microscopy, they were given a value of 0.5% to indicate their trace presences. When more than a single species belonged to a morphotype group in a single sample, the quantitative values from the morphological work were multiplied by the relative sequence abundance of each species in the load. All final compositions were standardized to a sum of 100%, by adding or subtracting the differences (induced by classifying records as 'trace') to all species with abundances > 1%.
